Qualifying Report

### 2019-01-16 Source/QualifyingReport/QualifyingReport.tex
QR edits 14-1-19

The potential impact of this research is immense. Research has shown that early diagnosis of people with AD or MCI improves sufferers quality of life and can, in some cases, slow the progress of the disease. Early diagnosis can increase the number of research opportunities for understanding the early stages of dementia and how the disease progresses so that more research can be conducted which may, in the future, lead to new treatments and other interventions.

@@

This report is structured in the following way. The context of the project, the aims and hypotheses of the research and the challenges the research presents are detailed in the remainder of this chapter. A literature review of current methods of cognitive assessment of MCI, an review of how language is said to be impacted in the MCI and Early AD population and a review of NLP and ML techniques that have the potential to be applied to this domain is presented in Chapter 2. The findings from a replication and extension of previous research using an existing dataset is presented in Chapter 3. I detail my future work plan including distinct tasks to be completed over the next two years in Chapter 4. Finally, my conclusions are summarised in Chapter 5.  
\section{Aims and Rationale}
The aim of my research is to find less burdensome ways of detecting dementia without the use of invasive procedures (taking bloods, or using medical equipment such as MRI's and EEG's) and without resorting to to time-consuming and expensive psychological tests. There is a lot of research into the analysis of language as a bio-marker for MCI and Early Dementia. Given that sampling a person's language is relatively effortless, my research looks at whether we can find bio-markers of MCI and Early Dementia in natural language.\newline
The broad question is can we use Machine Learning to assist in the early diagnosis of dementia and/or mild cognitive impairment through language produced naturally rather than language that is produced as part of a formal cognitive assessment. There have been a number of studies from the psychological perspective that look at the changes in language in those with Dementia. These studies have identified a number of speech and language differences between healthy controls and those with dementia. Some of these features have been validated in work using machine learning to classify healthy controls from those with dementia. However, these have been done in lab settings, with formal cognitive assessment tasks such as the 'picture description task'. While this is help to learn that machine learning can be used to differentiate between healthy controls and those with dementia, this doesn't take the burden off healthcare professionals who have to conduct the assessments and other professionals who may be used to transcribe these assessments.

@@

Alzheimer's Disease is a neurodegenerative disease in which, from a physiological perspective, the brain develops neurofibrillary tangles and neuritic plaques along with the deterioration and loss of cortical neurons and synapses. Whilst a definitive diagnosis of Dementia can only produced at post-mortem, there are a number of clinical indicators from psychological perspective that can indicate that dementia is present. From a clinical psychology perspective, those who have dementia have cognitive deficits such as problems with episodic and semantic memory, organising and planning, difficulties with language and visuospatial deficits\cite{McKhann2011}. In addition, these symptoms are often accompanied by emotional difficulties such as depression and irritability and behavioural difficulties. 

@@

Current thinking suggests that the cognitive deficits associated with AD often begin before the clinical symptoms of the disease become apparent. Researchers propose that neurofibrillary tangles and other associated physiological effects of AD develop over time and alter cognitive function until a threshold is reached and clinical symptoms become more obvious \cite{Nestor2006}. The case of Iris Murdoch illustrates this theory well. Le et al \cite{Le2011}found, in their analysis of three writers and the novels they wrote, that Iris Murdoch's work declined subtly over time, but there was a steep drop off in the use of language in her last novel. If this theory holds true more generally, it should be possible to detect subtle cognitive changes in language and memory before a clinical diagnosis can be formed. However, one of the challenges of this approach is differentiating natural cognitive decline due to aging with decline due to a form of dementia. Albert and his team have worked to define clinical criteria which professionals can use to diagnose MCI due to AD and differentiate this from age-associated memory impairment and age-associated cognitive decline. They note that the diagnosis of MCI requires evidence of intra-individual change and optimally requires evaluation at two or more points \cite{Albert2011}.

@@

One of the most common ways in which clinicians traditionally make an early diagnosis of cognitive impairment is through the use of the Mini Mental State Examination (MMSE) \cite{Folstein1975}. The MMSE is a brief questionnaire consisting of eleven questions which tests cognitive aspects of mental function and requires only 5-10 minutes to administer \cite{Folstein1975}. The MMSE is chosen due to it's effectiveness at assessing a person's cognitive mental state at a specific point in time, as well as being as sensitive to changes as a more detailed and complex assessment such as the Wechler Adult Intelligence Scale \cite{Folstein1975}. Whilst the MMSE is useful as a brief screening tool it has it's limitations. The MMSE was not specifically created to screen for dementias and therefore does not interrogate key aspects of cognitive impairment known to be affected in dementia. It also has limited value in assessing under-educated subjects and a meta-analysis on the effectiveness of the MMSE as a diagnostic tool for dementia to showed that it's accuracy was low (sensitivity between 78.4\% and 85.1\% and specificity between 81.3\% and 87.8\%). As the MMSE is shown to have low accuracy specifically in the diagnosis of dementia, it becomes necessary for professionals to employ the use of other tools or measures such as the Free Cued Selective Reminding Test (FCSRT) \cite{Grober2010} or the Montreal Cognitive Assessment (MoCA) \cite{Davis2015}. These tests have the benefits of being much more accurate at diagnosing cognitive impairment and discriminating between dementia and other types of cognitive impairment at the cost time and training of psychological professionals such as clinical psychologists in administering these tests. Given the burden on these professionals is likely to increase due to a growing elderly population or in the case of developing countries where the clinician / patient ratio is already unsustainable, it seems useful to find less burdensome ways to aid professionals in identifying those at risk of developing dementia.

@@

The two main ways in which diagnosis is performed is through assessment of memory and language. Tests of memory are classically among the most accurate ways of diagnosing dementia, however these tests suffer from the same reliance on trained clinicians to administer these tests in a clinical setting. Language however is a lot easier to collect and can be done in more naturalistic settings. As with memory, these tests can be done over time and would be able to chart a patients language degeneration over time. Given that language is less intrusive to test and requires a lot of the cognitive processes that may be impacted by AD, a lot of research has focused on measure decline in the use of language in those with AD. There are a number of difficulties to watch out for with this approach. There are a wide number of factors that are involved in language degeneration in the elderly, and consequently there will be an expected amount of variability between subjects. The administration of such tests may induce nervousness and discomfort which may impact performance, and also repeatedly administering the same language tests to tests for differences over time by be confounded by improved performance at tasks via practice effects. Also, generally speaking the language tests that are currently being used do not necessarily describe patients 'real performance' in language production. However, there is enough promise in this approach such that it could help further our understanding of the disease, it's progression and the parts of the brain affected in the early stages.

@@

Emery \cite{Emery2000} completed a literature review looking at all the potential language deficits that could exist in those with AD and / or MCI. She divided these deficits into four levels of language: Phonology, Morphology, Syntax and Semantics. She proposed that language and the processes involved in language are hierarchical in nature and that language moves from simple units of construction (Phonology and Morphology), and build layers of complexity and sophistication (Syntax and Semantics). She found that people with AD generally had intact Phonology and Morphology but more impaired Syntax and Semantics. She stated that the language forms we learn last are the first to deteriorate. We generally learn language in small simple units initially and build syntax and complexity as we are more comfortable with language.

@@

\subsection{Existing Neuropsychological Measures of Cognitive Impairment, and Repeatable Battery of Neuropsychological Tests}

@@

\subsubsection{Repeatable Battery for the Assessment of Neuropsychological Status}

@@

There has been a significant research in the area of language deterioration as a means of detecting Alzheimer's Disease. This usually takes the form analysis of speech recorded as part of a cognitive assessment such as the Picture Description Task \cite{Orimaye2008}. Given that language samples are the relatively easy to collect, research has moved towards analysis of spontaneous speech. An good example of this type of research is the study conducted by Berisha and Liss which looked at the differences in language use between two US presidents, Ronald Reagan (who would go on to receive a diagnosis of Dementia) and George H.W. Bush who acted as a matched control based on Age \citep{Berisha2015}. They found several differences in language use which they felt acted as indicators of Reagan's difficulties with language due to dementia. These significant differences were in the number of unique words used per speech, the use of non-specific nouns and fillers and low-imageability verbs \citep{Berisha2015}. 

@@

\chapter{Future Work}
\section{Introduction}
This section looks at my research aims and objectives given the literature review presented above. This section will then go on to explore any challenges to my research. I then present the work I have planned for the next two years, divided into distinct tasks with their own aims and objectives. As part of this I will identify target conferences where I could present my work. Finally, I look at potential risks and ethical concerns that may occur as part of my research.
\section{Research Aims and Objectives}
\par
The broad aim of my research is to explore whether the application of Natural Language Processing and Machine Learning techniques to spontaneous speech samples can improve our ability to diagnose Mild Cognitive Impairment in patients that will directly lead to Alzheimer's Disease. Another aim of my research is to provide some clarity as to which language features are most important to measure when looking for evidence of MCI and/or AD. Finally, an additional outcome of my research is the development of a repeatable battery of neuropsychological tests that specifically focus on being sensitive to Mild Cognitive Impairment.
\begin{itemize}
	\item \textbf{Hypothesis} Is monitoring language samples and other measures of cognitive decline over the course of the year enough to detect cognitive decline in a population of those with Subjective Cognitive Decline and Mild Cognitive Impairment.
	\item \textbf{Hypothesis} Those with MCI will show a different language profile to controls if matched similarly for age, gender, education and other significant factors.
	\item \textbf{Hypothesis} The new neuropsychological battery of tests will be more sensitive at detecting mild cognitive impairment than existing measures.
	\item \textbf{Hypothesis} Given a number of samples of language over a given time frame, it is possible to detect language decline in participants such that you can differentiate between normal language decline, those with Alzheimer's Disease and decline in those with MCI.
\end{itemize}

>>>>

The potential impact of this research in this area is immense. Research has shown that early diagnosis of people with AD or MCI improves sufferers quality of life and can, in some cases, slow the progress of the disease. Early diagnosis can increase the number of research opportunities for understanding the early stages of dementia and how the disease progresses so that more research can be conducted which may, in the future, lead to new treatments and other interventions.

@@

This report is structured in the following way. A literature review of current methods of cognitive assessment of MCI, an review of how language is said to be impacted in the MCI and Early AD population and a review of NLP and ML techniques that have the potential to be applied to this domain is presented in Chapter 2. The findings from a replication and extension of previous research using an existing dataset is presented in Chapter 3. I detail the aims of my research and my proposed future work plan including distinct tasks to be completed over the next two years in Chapter 4. Finally, my conclusions are summarised in Chapter 5.

@@

Alzheimer's Disease (AD) is a neurodegenerative disease in which, from a physiological perspective, the brain develops neurofibrillary tangles and neuritic plaques along with the deterioration and loss of cortical neurons and synapses. Whilst a definitive diagnosis of Dementia can only produced at post-mortem, there are a number of clinical indicators from psychological perspective that can indicate dementia is present. From a clinical psychology perspective, those who have dementia demonstrate cognitive deficits such as problems with episodic and semantic memory, organizing and planning, difficulties with language and visuospatial deficits\cite{McKhann2011}. In addition, these symptoms are often accompanied by emotional problems such as depression and behavioural difficulties. 

@@

Current thinking suggests that the cognitive deficits associated with AD often begin before the clinical symptoms of the disease become apparent. Researchers propose that neurofibrillary tangles and other associated physiological effects of AD develop over time and alter cognitive function until a threshold is reached and clinical symptoms become more obvious \cite{Nestor2006}. The case of Iris Murdoch, who had a confirmed diagnosis of dementia, illustrates this theory well. Le et al \cite{Le2011}found, in their analysis of three writers and the novels they wrote, that Iris Murdoch's work declined subtly over time, but there was a steep drop off in the use of language in her last novel when, it is theorized, the symptoms of AD manifested themselves more significantly. If this theory holds true more generally, it should be possible to detect subtle cognitive changes in language and memory before a clinical diagnosis can be formed. However, one of the challenges of this approach is differentiating natural cognitive decline due to aging with decline due to a form of dementia. Albert and his team have worked to define clinical criteria which professionals can use to diagnose MCI due to AD and differentiate this from age-associated memory impairment and age-associated cognitive decline. They note that the diagnosis of MCI requires evidence of intra-individual change and optimally requires evaluation at two or more points \cite{Albert2011}.

@@

One of the most common ways in which clinicians traditionally make an early diagnosis of cognitive impairment is through the use of the Mini Mental State Examination (MMSE) \cite{Folstein1975}. The MMSE is a brief questionnaire consisting of eleven questions which tests cognitive aspects of mental function and requires only 5-10 minutes to administer \cite{Folstein1975}. The MMSE is chosen due to it's effectiveness at assessing a person's cognitive mental state at a specific point in time, as well as being as sensitive to changes as a more detailed and complex assessment such as the Wechler Adult Intelligence Scale \cite{Folstein1975}. Whilst the MMSE is useful as a brief screening tool it has it's limitations. The MMSE was not specifically created to screen for dementias and therefore does not interrogate key aspects of cognitive impairment known to be affected in dementia. It also has limited value in assessing under-educated subjects and a meta-analysis on the effectiveness of the MMSE as a diagnostic tool for dementia showed that it's accuracy was low (sensitivity between 78.4\% and 85.1\% and specificity between 81.3\% and 87.8\%). As the MMSE is shown to have low accuracy specifically in the diagnosis of dementia, it becomes necessary for professionals to employ the use of other tools or measures such as the Free Cued Selective Reminding Test (FCSRT) \cite{Grober2010} or the Montreal Cognitive Assessment (MoCA) \cite{Davis2015}. These tests have the benefits of being much more accurate at diagnosing cognitive impairment and discriminating between dementia and other types of cognitive impairment at the cost time and training of psychological professionals such as clinical psychologists in administering these tests. Given the burden on these professionals is likely to increase due to a growing elderly population or in the case of developing countries where the clinician / patient ratio is already unsustainable, it seems useful to find less burdensome ways to aid professionals in identifying those at risk of developing dementia.

@@

The two main ways in which diagnosis is performed is through assessment of memory and language. Tests of memory are classically among the most accurate ways of diagnosing dementia, however these tests suffer from the same reliance on trained clinicians to administer these tests in a clinical setting. Language however is a lot easier to collect and can be done in more naturalistic settings. As with memory, these tests can be done over time and would be able to chart a patients language degeneration over time. Given that language is less intrusive to test and requires a lot of the cognitive processes that may be impacted by AD, a lot of research has focused on measure decline in the use of language in those with AD. There are a number of difficulties to watch out for with this approach. There are a wide number of factors that are involved in language degeneration in the elderly, and consequently there will be an expected amount of variability between subjects. The administration of such tests may induce nervousness and discomfort which may impact performance, and also repeatedly administering the same language tests for differences over time may be confounded by improved performance at tasks via practice effects. However, there is enough promise in this approach such that it could help further our understanding of the disease, it's progression and the parts of the brain affected in the early stages.

@@

Emery \cite{Emery2000} completed a literature review looking at all the potential language deficits that could exist in those with AD and / or MCI. She divided these deficits into four levels of language: Phonology, Morphology, Syntax and Semantics. She proposed that language and the processes involved in language are hierarchical in nature and that language moves from simple units of construction (Phonology and Morphology), and build layers of complexity and sophistication (Syntax and Semantics). She found that people with AD generally had intact Phonology and Morphology but more impaired Syntax and Semantics. She asserted that the language forms we learn last are the first to deteriorate as we generally learn language in small simple units initially and build syntax and complexity as we are more comfortable with language.

@@

\section{Existing Neuropsychological Measures of Cognitive Impairment, and Repeatable Battery of Neuropsychological Tests}

@@

\subsection{Repeatable Battery for the Assessment of Neuropsychological Status}

@@

There has been a significant research in the area of language deterioration as a means of detecting Alzheimer's Disease. This usually takes the form analysis of speech recorded as part of a cognitive assessment such as the Picture Description Task \cite{Orimaye2008}. Given that language samples are the relatively easy to collect, research has moved towards analysis of spontaneous speech. An good example of this type of research is the study conducted by Berisha and Liss which looked at the differences in language use between two US presidents, Ronald Reagan (who would go on to receive a diagnosis of Dementia) and George H.W. Bush who acted as a matched control based on Age \cite{Berisha2015}. They found several differences in language use which they felt acted as indicators of Reagan's difficulties with language due to dementia. These significant differences were in the number of unique words used per speech, the use of non-specific nouns and fillers and low-imageability verbs \cite{Berisha2015}. 

@@

\chapter{Future Work}
This section looks at my research aims and objectives given the literature review presented above. This section will then go on to explore any challenges to my research. I then present the work I have planned for the next two years, divided into distinct tasks with their own aims and objectives. As part of this I will identify target conferences where I could present my work. Finally, I look at potential risks and ethical concerns that may occur as part of my research.
\section{Research Aims and Objectives}
\section{Research Aims}
The aim of my research is to find less burdensome ways of detecting dementia without the use of invasive procedures (taking bloods, or using medical equipment such as MRI's and EEG's) and without resorting to time-consuming and expensive psychological tests. There is a lot of research into the analysis of language as a marker for MCI and Early Dementia \cite{Taler2008, Boschi2017} and there have been calls for research that explores the use of Natural Language Processing and Machine Learning techniques to aid in the analysis of language \cite{Boschi2017}. It is this area which is the focus of my research.
Given that sampling a person's language is relatively effortless, my proposed research will look at whether we can find markers of MCI in natural language. I focus particularly on MCI as this is often a precursor to AD and therefore detection of MCI in an elderly population would allow I plan to explore the use Natural Language Processing and Machine Learning techniques to assist in this process, and focus on MCI specifically. early diagnosis of dementia and/or mild cognitive impairment through language produced naturally rather than language that is produced as part of a formal cognitive assessment. There have been a number of studies from the psychological perspective that look at the changes in language in those with Dementia. These studies have identified a number of speech and language differences between healthy controls and those with dementia. Some of these features have been validated in work using machine learning to classify healthy controls from those with dementia. However, these have been done in lab settings, with formal cognitive assessment tasks such as the 'picture description task'. While this is help to learn that machine learning can be used to differentiate between healthy controls and those with dementia, this doesn't take the burden off healthcare professionals who have to conduct the assessments and other professionals who may be used to transcribe these assessments.
The broad aim of my research is to explore whether the application of Natural Language Processing and Machine Learning techniques to spontaneous speech samples can improve our ability to diagnose Mild Cognitive Impairment in patients that will directly lead to Alzheimer's Disease. Another aim of my research is to provide some clarity as to which language features are most important to measure when looking for evidence of MCI and/or AD. Finally, an additional outcome of my research is the development of a repeatable battery of neuropsychological tests that specifically focus on being sensitive to Mild Cognitive Impairment.
\begin{itemize}
	\item \textbf{Hypothesis} Is monitoring language samples and other measures of cognitive decline over the course of the year enough to detect cognitive decline in a population of those with Subjective Cognitive Decline and Mild Cognitive Impairment.
	\item \textbf{Hypothesis} Those with MCI will show a different language profile to controls if matched similarly for age, gender, education and other significant factors.
	\item \textbf{Hypothesis} The new neuropsychological battery of tests will be more sensitive at detecting mild cognitive impairment than existing measures.
	\item \textbf{Hypothesis} Given a number of samples of language over a given time frame, it is possible to detect language decline in participants such that you can differentiate between normal language decline, those with Alzheimer's Disease and decline in those with MCI.
\end{itemize}
